---
bibliography: references.bib
---

# Introduction {#intro}

This web page documents the analysis for for a paper by 
[Marte Fandrem](https://github.com/martef)
on the effects of hydrological disturbance on the growth of Sphagnum
mosses on a Norwegian bog.

The bog is a relatively large ombrotrophic bog previously used for peat 
extraction, and also drained and used for forest plantations. 

Sphagnum growth was measured using the cranked wire technique on permanent
vegetation quadrats (n = 28) from 2017 to 2022 (5 growing seasons). 
Each vegetation quadrat has 16 points (i.e. pins) and each pin was measured up 
to 4 times on each sampling occasion (on each side and by different people).
The first 18 vegetation plots were initiated in 2017 and are paired with 
water table well that log the water table depth. 
These vegetation plots are situated in homogeneous vegetation types (bog lawns),
but with varying distance to disturbances (extraction sites and ditches).
The quadrats numbered 18-28 were initiated in 2021 and a placed relatively
for from the hydrological disturbances, but in varying communities (hollows and
hummoks).

```{r setup}
#| code-summary: "Load packages"
library(tidyverse)
library(sf)
library(tmap)
```


```{r mapData}
#| code-summary: "Import data"
path    <- "data/shapeFiles/"
ditches <- sf::read_sf(paste0(path, "ditches/grofter.shp"))
massifs  <- sf::read_sf(paste0(path, "massifs/hostadmyra_myrmassiv.shp")) 

#extrSites <- sf::read_sf(paste0(path, "peatExtractionSites/Torvtak.shp"))
# This data is the same as the masifs.

quadrats <- sf::read_sf(paste0(path, "vegetationQuadrats1/veipunkter_vannbronner_ruter_2017.shp")) |>
  bind_rows(sf::read_sf(paste0(path, "vegetationQuadrats2/vegetasjonsruter_19-30.shp")))
```

```{r studyMap}
#| fig.cap: "Map of study area. Green circles are permanent vegetation quadrats.
#| The shades of grey indicator different bog massifs."
#| code-summary: "Create map"
#| cache: true
tm_shape(massifs) +
  tm_polygons(col = "Name",
              palette = c("grey60", "grey70", "grey80"),
              legend.show=F) +
  tm_shape(ditches |>
             mutate("Ditch age" = case_when(
               Name == "Eldre" ~ "Old",
               Name == "Nyere" ~ "More recent",
               .default = "Old"
             ))) +
  tm_lines(col = "Ditch age",
           lwd=2,
           palette = c("red", "orange")) +
  #tm_shape(extrSites) +
  #tm_polygons()
  tm_shape(quadrats) +
  tm_dots(size = 2,
          shape=21,
          col = "green",
          alpha = 0.5)  +
  tm_scale_bar(position = c("left", "bottom")) +
  tm_compass()
```

::: {.callout.note}
The shape file with the position of vegetation qadrats do not contain the
qadrat ID for all cases. We need to add that.
:::
